---
title: "Preliminary model robustness checks"
subtitle: "Cumulative mortality and forest-community climate-niche change"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(data.table)
library(knitr)

run_dir <- Sys.getenv("ANALYSIS_RUN_DIR")
robust_dir <- file.path(run_dir, "robustness")
coefficients <- fread(file.path(robust_dir, "coefficients.csv"))
fits <- fread(file.path(robust_dir, "model_fit.csv"))
common <- fread(file.path(robust_dir, "common_history_summary.csv"))
support <- fread(file.path(robust_dir, "condition_support_summary.csv"))
stability <- fread(file.path(robust_dir, "mortality_stability.csv"))
interactions <- fread(file.path(robust_dir, "life_stage_interaction_tests.csv"))
pairwise <- fread(file.path(robust_dir, "life_stage_pairwise_differences.csv"))
baseline_check <- fread(file.path(robust_dir, "baseline_reproduction.csv"))

term_labels <- c(
  fire_cumulative_mortality_pct = "Fire mortality",
  insect_cumulative_mortality_pct = "Insect mortality",
  disease_cumulative_mortality_pct = "Disease mortality",
  cumulative_site_CWD_mm = "Cumulative site CWD",
  mean_monthly_site_CWD_mm = "Mean monthly site CWD",
  full_survey_period_years = "Survey period"
)
response_labels <- c(
  temperature = "Temperature CWM",
  precipitation = "Precipitation CWM",
  CWD = "CWD CWM"
)
fmt <- function(x, digits = 4) formatC(x, digits = digits, format = "g")
```

## What these checks answer

The approved preliminary model remains the reference model. These checks ask
whether its disturbance coefficients depend on:

1. comparing different condition histories across life stages;
2. retaining conditions that occupy only part of a plot;
3. broad geographic differences among states;
4. representing site water deficit as cumulative exposure while also including
   survey duration.

They also test life-stage differences directly on identical histories. They do
not introduce a new severity definition or replace cumulative mortality.

## Main findings

```{r finding-values, include=FALSE}
supported <- stability[p_value_baseline < 0.05]
supported_stage <- supported[group != "community"]
n_interactions <- interactions[p_value < 0.05, .N]
state_max_r2 <- fits[scenario == "state_fixed_effects", max(r_squared)]
```

- All `r nrow(supported)` disturbance coefficients whose baseline 95% interval
  excluded zero retained the same direction across every applicable check.
- All `r nrow(supported)` remained below p = 0.05 with state effects, mean
  monthly CWD, and cumulative CWD without the separate survey-period term.
- `r supported[p_value_condition_prop_80 < 0.05, .N]` of those
  `r nrow(supported)` remained below p = 0.05 when conditions were required to
  occupy at least 80% of the plot.
- On the smaller identical-history sample,
  `r supported_stage[p_value_common_histories < 0.05, .N]` of the
  `r nrow(supported_stage)` supported life-stage coefficients remained below
  p = 0.05. Their loss of precision should not be read as proof of no effect.
- `r n_interactions` of the `r nrow(interactions)` joint life-stage tests were
  below p = 0.05. The seedling contrasts therefore cannot be explained only by
  different sample sizes.
- Even the state-effects models explain at most
  `r round(100 * state_max_r2, 2)`% of response variance. These remain weakly
  predictive, observational preliminary models.

## Reproduction check

The robustness script refit the approved models before changing anything. The
largest absolute coefficient difference was
`r format(max(baseline_check$estimate_absolute_difference), scientific = TRUE)`
and the largest standard-error difference was
`r format(max(baseline_check$standard_error_absolute_difference), scientific = TRUE)`.

## Are the life-stage samples different?

```{r common-history-table}
common_show <- copy(common)
common_show[, response := response_labels[response]]
setnames(
  common_show,
  c(
    "response", "common_histories", "seedling_baseline_histories",
    "sapling_baseline_histories", "adult_baseline_histories"
  ),
  c(
    "Response", "Histories shared by all stages", "Seedling histories",
    "Sapling histories", "Adult histories"
  )
)
kable(common_show, format.args = list(big.mark = ","))
```

The common-history models remove sample-composition differences: seedlings,
saplings, and adults are fit on the same condition histories for each response.

## Do life-stage coefficients actually differ?

```{r interaction-table}
interaction_show <- copy(interactions)
interaction_show[, `:=`(
  response = response_labels[response],
  predictor = term_labels[predictor],
  p_value = fmt(p_value, 3)
)]
setnames(
  interaction_show,
  c("response", "predictor", "histories", "statistic", "df", "p_value"),
  c("Response", "Predictor", "Common histories", "Wald statistic", "df", "p")
)
kable(interaction_show, digits = 3, format.args = list(big.mark = ","))
```

The joint test asks whether the seedling, sapling, and adult coefficients are
equal. A small p-value means the life-stage contrast is larger than expected
from sampling uncertainty under this model.

```{r pairwise-table}
pair_show <- copy(pairwise[
  comparison %in% c("Saplings - Seedlings", "Adults - Seedlings")
])
pair_show[, `:=`(
  response = response_labels[response],
  predictor = term_labels[predictor],
  estimate = fmt(estimate),
  conf_low = fmt(conf_low),
  conf_high = fmt(conf_high),
  p_value = fmt(p_value, 3)
)]
setnames(
  pair_show,
  c(
    "response", "predictor", "comparison", "estimate", "conf_low",
    "conf_high", "p_value"
  ),
  c("Response", "Predictor", "Comparison", "Difference", "95% low", "95% high", "p")
)
kable(pair_show[, -"std_error"], align = "lllrrrr")
```

## Does condition spatial support matter?

The primary analysis requires a condition to occupy at least 30% of the plot.
The 80% check retains histories whose condition remains at least 80% of the plot
through every interval. This reduces, but cannot eliminate, concern that a
condition-level disturbance may not represent the exact microplot where
seedlings were tallied.

```{r support-table}
support_show <- dcast(
  support,
  group_label ~ threshold,
  value.var = "histories"
)
setnames(support_show, "group_label", "Community group")
kable(support_show, format.args = list(big.mark = ","))
```

## Are the disturbance results stable?

```{r stability-figures, out.width='100%'}
for (slug in c("temperature", "precipitation", "cwd")) {
  print(knitr::include_graphics(file.path(
    robust_dir, "figures",
    paste0(slug, "_mortality_coefficient_stability.png")
  )))
}
```

```{r sign-stability-table}
sign_show <- stability[, .(
  coefficients = .N,
  sign_consistent = sum(sign_consistent),
  sign_changes = sum(!sign_consistent)
), by = .(response, group_label)]
sign_show[, response := response_labels[response]]
setnames(
  sign_show,
  c("response", "group_label", "coefficients", "sign_consistent", "sign_changes"),
  c("Response", "Community group", "Disturbance coefficients", "Stable sign", "Sign changes")
)
kable(sign_show)
```

## Cumulative CWD and survey duration

Three formulations are compared in the stability figure:

- the approved model with cumulative CWD and survey duration;
- mean monthly CWD and survey duration;
- cumulative CWD without a separate duration term.

The alternative formulations are diagnostics. Cumulative CWD remains the
intended total-exposure predictor unless the collaborators choose otherwise.

## Interpretation

These are observational preliminary models with very low explained variance.
Coefficient direction and uncertainty are more informative here than p-values
alone. A stable coefficient is not automatically causal, and an unstable
coefficient identifies a modeling or sampling dependency that should be
resolved before final inference.
