---
title: "Preliminary cumulative-mortality model results"
subtitle: "Life-stage and pooled climate-niche change models"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  out.width = "100%"
)

run_dir <- Sys.getenv("ANALYSIS_RUN_DIR")
if (!nzchar(run_dir) || !dir.exists(run_dir)) {
  stop("ANALYSIS_RUN_DIR must identify the model-run output directory.")
}
run_dir <- normalizePath(run_dir, winslash = "/")
project_dir <- normalizePath(".", winslash = "/")
table_dir <- file.path(run_dir, "tables")
figure_dir <- file.path(run_dir, "figures")
output_dir <- run_dir

response_order <- c("temperature", "precipitation", "cwd")
response_labels <- c(
  temperature = "Temperature CWM change",
  precipitation = "Precipitation CWM change",
  cwd = "CWD CWM change"
)
section_order <- c("raw_relationship", "marginal_effect")
section_labels <- c(
  raw_relationship = "Raw relationships",
  marginal_effect = "Adjusted marginal effects"
)
driver_order <- c(
  "Fire mortality",
  "Insect mortality",
  "Disease mortality",
  "Cumulative site CWD"
)

table_style_path <- file.path(table_dir, "sjplot_table_style.html")
table_fragment_paths <- setNames(
  file.path(table_dir, paste0(response_order, "_models_fragment.html")),
  response_order
)
figure_manifest_path <- file.path(figure_dir, "figure_manifest.csv")
sample_path <- file.path(output_dir, "sample_flow.csv")

required_assets <- c(
  table_style_path,
  unname(table_fragment_paths),
  figure_manifest_path,
  sample_path
)
missing_assets <- required_assets[!file.exists(required_assets)]
if (length(missing_assets) > 0L) {
  stop(
    "The report assets have not been fully produced:\n- ",
    paste(missing_assets, collapse = "\n- ")
  )
}

read_html_fragment <- function(path) {
  paste(readLines(path, warn = FALSE, encoding = "UTF-8"), collapse = "\n")
}

emit_html_fragment <- function(path) {
  cat(as.character(htmltools::HTML(read_html_fragment(path))), "\n\n")
}

resolve_figure_path <- function(path) {
  is_absolute <- grepl("^([A-Za-z]:[/\\\\]|/)", path)
  candidates <- if (is_absolute) {
    path
  } else {
    c(
      file.path(run_dir, path),
      file.path(figure_dir, path),
      file.path(project_dir, path),
      path
    )
  }
  existing <- candidates[file.exists(candidates)]
  if (length(existing) == 0L) {
    return(NA_character_)
  }
  normalizePath(existing[1L], winslash = "/")
}

samples <- read.csv(sample_path, check.names = FALSE, stringsAsFactors = FALSE)
figure_manifest <- read.csv(
  figure_manifest_path,
  check.names = FALSE,
  stringsAsFactors = FALSE
)

required_manifest_columns <- c(
  "section", "response", "driver_label", "figure_path", "caption", "plot_order"
)
missing_manifest_columns <- setdiff(required_manifest_columns, names(figure_manifest))
if (length(missing_manifest_columns) > 0L) {
  stop(
    "figure_manifest.csv is missing required columns: ",
    paste(missing_manifest_columns, collapse = ", ")
  )
}

if (!setequal(unique(figure_manifest$section), section_order)) {
  stop("Manifest section values must be raw_relationship and marginal_effect.")
}
if (!setequal(unique(figure_manifest$response), response_order)) {
  stop("Manifest response values must be temperature, precipitation, and cwd.")
}
if (!setequal(unique(figure_manifest$driver_label), driver_order)) {
  stop(
    "Manifest driver_label values must be: ",
    paste(driver_order, collapse = "; ")
  )
}

expected_grid <- expand.grid(
  section = section_order,
  response = response_order,
  driver_label = driver_order,
  stringsAsFactors = FALSE
)
manifest_keys <- paste(
  figure_manifest$section,
  figure_manifest$response,
  figure_manifest$driver_label,
  sep = "__"
)
expected_keys <- paste(
  expected_grid$section,
  expected_grid$response,
  expected_grid$driver_label,
  sep = "__"
)
if (nrow(figure_manifest) != 24L || anyDuplicated(manifest_keys) ||
    !setequal(manifest_keys, expected_keys)) {
  stop(
    "figure_manifest.csv must contain exactly one plot for each of the 24 ",
    "section x response x driver combinations."
  )
}

if (anyNA(suppressWarnings(as.numeric(figure_manifest$plot_order))) ||
    anyDuplicated(figure_manifest$plot_order)) {
  stop("plot_order must contain 24 unique numeric values.")
}

figure_manifest$figure_path_resolved <- vapply(
  figure_manifest$figure_path,
  resolve_figure_path,
  character(1)
)
missing_figures <- figure_manifest$figure_path[is.na(figure_manifest$figure_path_resolved)]
if (length(missing_figures) > 0L) {
  stop(
    "The following manifest figures do not exist:\n- ",
    paste(missing_figures, collapse = "\n- ")
  )
}

figure_manifest$response <- factor(
  figure_manifest$response,
  levels = response_order
)
figure_manifest$section <- factor(
  figure_manifest$section,
  levels = section_order
)
figure_manifest$driver_label <- factor(
  figure_manifest$driver_label,
  levels = driver_order
)
figure_manifest$plot_order <- as.numeric(figure_manifest$plot_order)
figure_manifest <- figure_manifest[
  order(
    figure_manifest$response,
    figure_manifest$section,
    figure_manifest$driver_label,
    figure_manifest$plot_order
  ),
]
```

```{r report-style, results='asis'}
emit_html_fragment(table_style_path)
cat(
  "<style>",
  ".report-note {max-width: 1050px; line-height: 1.55;}",
  ".figure-caption {max-width: 1050px; margin: -0.5rem auto 2rem auto;",
  "color: #444; font-size: 0.95rem; line-height: 1.4;}",
  ".model-table-block {overflow-x: auto; margin-bottom: 2rem;}",
  "</style>\n"
)
```

# Overview

<div class="report-note">

This report assembles the current preliminary cumulative-mortality results for
seedling, sapling, adult, and pooled all-live forest-community climate niches.
The three responses are raw first-to-last changes in temperature,
precipitation, and climatic water deficit (CWD) community-weighted means.

Every model includes cumulative fire, insect, and disease mortality,
cumulative site-level TerraClimate CWD over the complete survey period, and
the length of that survey period. Model coefficients are unstandardized. The
model tables report HC1 uncertainty clustered by stable FIA plot.

For each response, the raw-relationship figures show the observed response and
one focal driver without covariate adjustment. The marginal-effect figures show
the fitted relationship for that driver at one shared covariate reference:
other agent mortalities equal zero, with site CWD and survey duration fixed at
their pooled-history medians when they are not focal. These are preliminary
associations, not final causal estimates.

</div>

# Model samples

```{r sample-table}
knitr::kable(
  samples,
  format = "html",
  caption = "Rows and stable FIA plots represented in the preliminary models.",
  big.mark = ",",
  table.attr = "style='width:100%; max-width:1050px;'"
)
```

```{r response-sections, results='asis', fig.show='asis'}
for (response_name in response_order) {
  cat("\n# ", response_labels[[response_name]], "\n\n", sep = "")
  cat("\n## Model results\n\n")
  cat("<div class='model-table-block'>\n")
  emit_html_fragment(table_fragment_paths[[response_name]])
  cat("</div>\n\n")

  for (section_name in section_order) {
    cat("\n## ", section_labels[[section_name]], "\n\n", sep = "")
    section_rows <- figure_manifest[
      figure_manifest$response == response_name &
        figure_manifest$section == section_name,
    ]

    for (driver_name in driver_order) {
      figure_row <- section_rows[
        as.character(section_rows$driver_label) == driver_name,
      ]
      cat("\n### ", driver_name, "\n\n", sep = "")
      image_path <- as.character(knitr::include_graphics(
        figure_row$figure_path_resolved,
        rel_path = FALSE
      ))
      # Objects printed inside a for-loop bypass knitr's top-level image hook.
      # Emit the validated path as Markdown so Pandoc embeds it in the
      # self-contained HTML instead of printing the path as ordinary text.
      cat("![", driver_name, "](", image_path, "){width=100%}\n\n", sep = "")
      cat(
        "\n<div class='figure-caption'>",
        htmltools::htmlEscape(figure_row$caption),
        "</div>\n\n",
        sep = ""
      )
    }
  }
}
```

# Interpretation guardrails

- Mortality predictors are cumulative percentages over each complete condition
  history; they are not annualized rates in these preliminary models.
- Cumulative site CWD is TerraClimate `def` exposure at the FIA site, not the
  CWD community-niche response.
- The raw figures are descriptive. Use the model tables and marginal-effect
  figures when discussing covariate-adjusted model results.
- Seedlings, saplings, adults, and the exploratory pooled all-live forest are
  shown as four separate model columns and plot panels.
