---
title: "Preliminary cumulative-mortality model results"
subtitle: "Individual and combined climate-niche change models"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  out.width = "100%"
)

run_dir <- Sys.getenv("ANALYSIS_RUN_DIR")
if (!nzchar(run_dir) || !dir.exists(run_dir)) {
  stop("ANALYSIS_RUN_DIR must identify the model-run output directory.")
}
run_dir <- normalizePath(run_dir, winslash = "/")
project_dir <- normalizePath(".", winslash = "/")
table_dir <- file.path(run_dir, "tables")
figure_dir <- file.path(run_dir, "figures")
output_dir <- run_dir

response_order <- c("temperature", "precipitation", "cwd")
response_labels <- c(
  temperature = "Temperature CWM change",
  precipitation = "Precipitation CWM change",
  cwd = "CWD CWM change"
)
section_order <- c("raw_relationship", "marginal_effect")
section_labels <- c(
  raw_relationship = "Raw relationships",
  marginal_effect = "Adjusted marginal effects"
)
driver_order <- c(
  "Fire mortality",
  "Insect mortality",
  "Disease mortality",
  "Cumulative site CWD"
)

table_style_path <- file.path(table_dir, "sjplot_table_style.html")
table_fragment_paths <- setNames(
  file.path(table_dir, paste0(response_order, "_models_fragment.html")),
  response_order
)
figure_manifest_path <- file.path(figure_dir, "figure_manifest.csv")
sample_path <- file.path(output_dir, "sample_flow.csv")

required_assets <- c(
  table_style_path,
  unname(table_fragment_paths),
  figure_manifest_path,
  sample_path
)
missing_assets <- required_assets[!file.exists(required_assets)]
if (length(missing_assets) > 0L) {
  stop(
    "The report assets have not been fully produced:\n- ",
    paste(missing_assets, collapse = "\n- ")
  )
}

read_html_fragment <- function(path) {
  paste(readLines(path, warn = FALSE, encoding = "UTF-8"), collapse = "\n")
}

emit_html_fragment <- function(path) {
  cat(as.character(htmltools::HTML(read_html_fragment(path))), "\n\n")
}

resolve_figure_path <- function(path) {
  is_absolute <- grepl("^([A-Za-z]:[/\\\\]|/)", path)
  candidates <- if (is_absolute) {
    path
  } else {
    c(
      file.path(run_dir, path),
      file.path(figure_dir, path),
      file.path(project_dir, path),
      path
    )
  }
  existing <- candidates[file.exists(candidates)]
  if (length(existing) == 0L) {
    return(NA_character_)
  }
  normalizePath(existing[1L], winslash = "/")
}

samples <- read.csv(sample_path, check.names = FALSE, stringsAsFactors = FALSE)
figure_manifest <- read.csv(
  figure_manifest_path,
  check.names = FALSE,
  stringsAsFactors = FALSE
)

required_manifest_columns <- c(
  "section", "response", "driver_label", "figure_path", "caption", "plot_order"
)
missing_manifest_columns <- setdiff(required_manifest_columns, names(figure_manifest))
if (length(missing_manifest_columns) > 0L) {
  stop(
    "figure_manifest.csv is missing required columns: ",
    paste(missing_manifest_columns, collapse = ", ")
  )
}

if (!setequal(unique(figure_manifest$section), section_order)) {
  stop("Manifest section values must be raw_relationship and marginal_effect.")
}
if (!setequal(unique(figure_manifest$response), response_order)) {
  stop("Manifest response values must be temperature, precipitation, and cwd.")
}
if (!setequal(unique(figure_manifest$driver_label), driver_order)) {
  stop(
    "Manifest driver_label values must be: ",
    paste(driver_order, collapse = "; ")
  )
}

expected_grid <- expand.grid(
  section = section_order,
  response = response_order,
  driver_label = driver_order,
  stringsAsFactors = FALSE
)
manifest_keys <- paste(
  figure_manifest$section,
  figure_manifest$response,
  figure_manifest$driver_label,
  sep = "__"
)
expected_keys <- paste(
  expected_grid$section,
  expected_grid$response,
  expected_grid$driver_label,
  sep = "__"
)
if (nrow(figure_manifest) != 24L || anyDuplicated(manifest_keys) ||
    !setequal(manifest_keys, expected_keys)) {
  stop(
    "figure_manifest.csv must contain exactly one plot for each of the 24 ",
    "section x response x driver combinations."
  )
}

if (anyNA(suppressWarnings(as.numeric(figure_manifest$plot_order))) ||
    anyDuplicated(figure_manifest$plot_order)) {
  stop("plot_order must contain 24 unique numeric values.")
}

figure_manifest$figure_path_resolved <- vapply(
  figure_manifest$figure_path,
  resolve_figure_path,
  character(1)
)
missing_figures <- figure_manifest$figure_path[is.na(figure_manifest$figure_path_resolved)]
if (length(missing_figures) > 0L) {
  stop(
    "The following manifest figures do not exist:\n- ",
    paste(missing_figures, collapse = "\n- ")
  )
}

figure_manifest$response <- factor(
  figure_manifest$response,
  levels = response_order
)
figure_manifest$section <- factor(
  figure_manifest$section,
  levels = section_order
)
figure_manifest$driver_label <- factor(
  figure_manifest$driver_label,
  levels = driver_order
)
figure_manifest$plot_order <- as.numeric(figure_manifest$plot_order)
figure_manifest <- figure_manifest[
  order(
    figure_manifest$response,
    figure_manifest$section,
    figure_manifest$driver_label,
    figure_manifest$plot_order
  ),
]
```

```{r report-style, results='asis'}
emit_html_fragment(table_style_path)
cat(
  "<style>",
  ".report-note {max-width: 1050px; line-height: 1.55;}",
  ".figure-caption {max-width: 1050px; margin: -0.5rem auto 2rem auto;",
  "color: #444; font-size: 0.95rem; line-height: 1.4;}",
  ".model-table-block {overflow-x: auto; margin-bottom: 2rem;}",
  ".zoomable-plot {cursor: zoom-in;}",
  ".plot-viewer {width: 96vw; max-width: none; height: 92vh; padding: 0;",
  "border: 1px solid #777;}",
  ".plot-viewer::backdrop {background: rgba(0, 0, 0, 0.75);}",
  ".plot-viewer-toolbar {display: flex; gap: 0.5rem; align-items: center;",
  "padding: 0.5rem 0.75rem; border-bottom: 1px solid #aaa; background: white;}",
  ".plot-viewer-toolbar button {min-width: 2.5rem;}",
  ".plot-viewer-help {margin-left: auto; color: #444;}",
  ".plot-viewer-viewport {height: calc(92vh - 3.25rem); overflow: hidden;",
  "display: flex; align-items: center; justify-content: center;",
  "background: white; touch-action: none; cursor: grab;}",
  ".plot-viewer-viewport.is-dragging {cursor: grabbing;}",
  ".plot-viewer-image {max-width: 100%; max-height: 100%;",
  "transform-origin: center center; user-select: none; -webkit-user-drag: none;}",
  "</style>\n"
)
```

# Overview

<div class="report-note">

This report assembles the current preliminary cumulative-mortality results for sapling, adult, and combined sapling-plus-adult forest-community climate niches. The three responses are raw first-to-last changes in temperature, precipitation, and climatic water deficit (CWD) community-weighted means.

Every model includes cumulative fire, insect, and disease mortality, cumulative site-level TerraClimate CWD over the complete survey period, and the length of that survey period. Model coefficients are unstandardized. The model tables report HC1 uncertainty clustered by stable FIA plot.

For each response, the raw-relationship figures show the observed response and one focal driver without covariate adjustment. The marginal-effect figures show the fitted relationship for that driver at one shared covariate reference: other agent mortalities equal zero, with site CWD and survey duration fixed at their combined-history medians when they are not focal. These are preliminary associations, not final causal estimates.

Select any plot to open a full-screen viewer. Use the mouse wheel or the controls to zoom, and drag the enlarged plot to pan. The displayed data are not cropped.

</div>

# Model samples

```{r sample-table}
knitr::kable(
  samples,
  format = "html",
  caption = "Rows and stable FIA plots represented in the preliminary models.",
  big.mark = ",",
  table.attr = "style='width:100%; max-width:1050px;'"
)
```

```{r response-sections, results='asis', fig.show='asis'}
for (response_name in response_order) {
  cat("\n# ", response_labels[[response_name]], "\n\n", sep = "")
  cat("\n## Model results\n\n")
  cat("<div class='model-table-block'>\n")
  emit_html_fragment(table_fragment_paths[[response_name]])
  cat("</div>\n\n")

  for (section_name in section_order) {
    cat("\n## ", section_labels[[section_name]], "\n\n", sep = "")
    section_rows <- figure_manifest[
      figure_manifest$response == response_name &
        figure_manifest$section == section_name,
    ]

    for (driver_name in driver_order) {
      figure_row <- section_rows[
        as.character(section_rows$driver_label) == driver_name,
      ]
      cat("\n### ", driver_name, "\n\n", sep = "")
      image_path <- as.character(knitr::include_graphics(
        figure_row$figure_path_resolved,
        rel_path = FALSE
      ))
      # Objects printed inside a for-loop bypass knitr's top-level image hook.
      # Emit the validated path as Markdown so Pandoc embeds it in the
      # self-contained HTML instead of printing the path as ordinary text.
      cat(
        "![", driver_name, "](", image_path,
        "){.zoomable-plot width=100%}\n\n",
        sep = ""
      )
      cat(
        "\n<div class='figure-caption'>",
        htmltools::htmlEscape(figure_row$caption),
        "</div>\n\n",
        sep = ""
      )
    }
  }
}
```

<dialog id="plot-viewer" class="plot-viewer" aria-label="Zoomed plot viewer">
<div class="plot-viewer-toolbar">
<button type="button" id="plot-viewer-minus" aria-label="Zoom out">−</button>
<button type="button" id="plot-viewer-reset">Reset</button>
<button type="button" id="plot-viewer-plus" aria-label="Zoom in">+</button>
<span class="plot-viewer-help">Wheel to zoom; drag to pan</span>
<button type="button" id="plot-viewer-close">Close</button>
</div>
<div id="plot-viewer-viewport" class="plot-viewer-viewport">
<img id="plot-viewer-image" class="plot-viewer-image" alt="">
</div>
</dialog>

<script>
document.addEventListener("DOMContentLoaded", function () { const dialog = document.getElementById("plot-viewer"); const viewport = document.getElementById("plot-viewer-viewport"); const viewerImage = document.getElementById("plot-viewer-image"); const closeButton = document.getElementById("plot-viewer-close"); const resetButton = document.getElementById("plot-viewer-reset"); const plusButton = document.getElementById("plot-viewer-plus"); const minusButton = document.getElementById("plot-viewer-minus"); let scale = 1; let offsetX = 0; let offsetY = 0; let dragging = false; let previousX = 0; let previousY = 0;

function applyTransform() {
    viewerImage.style.transform =
      `translate(${offsetX}px, ${offsetY}px) scale(${scale})`;
}

function resetView() {
    scale = 1;
    offsetX = 0;
    offsetY = 0;
    applyTransform();
}

function changeScale(multiplier) {
    scale = Math.min(8, Math.max(1, scale * multiplier));
    if (scale === 1) {
      offsetX = 0;
      offsetY = 0;
    }
    applyTransform();
}

document.querySelectorAll("img.zoomable-plot").forEach(function (image) {
    image.setAttribute("tabindex", "0");
    image.setAttribute("role", "button");
    image.setAttribute("aria-label", `${image.alt}. Open zoomed plot.`);

    function openViewer() {
      viewerImage.src = image.src;
      viewerImage.alt = image.alt;
      resetView();
      dialog.showModal();
    }

    image.addEventListener("click", openViewer);
    image.addEventListener("keydown", function (event) {
      if (event.key === "Enter" || event.key === " ") {
        event.preventDefault();
        openViewer();
      }
    });
});

viewport.addEventListener("wheel", function (event) {
    event.preventDefault();
    changeScale(event.deltaY < 0 ? 1.2 : 1 / 1.2);
}, { passive: false });

viewport.addEventListener("pointerdown", function (event) {
    if (scale <= 1) return;
    dragging = true;
    previousX = event.clientX;
    previousY = event.clientY;
    viewport.classList.add("is-dragging");
    viewport.setPointerCapture(event.pointerId);
});

viewport.addEventListener("pointermove", function (event) {
    if (!dragging) return;
    offsetX += event.clientX - previousX;
    offsetY += event.clientY - previousY;
    previousX = event.clientX;
    previousY = event.clientY;
    applyTransform();
});

function stopDragging() {
    dragging = false;
    viewport.classList.remove("is-dragging");
}

viewport.addEventListener("pointerup", stopDragging); viewport.addEventListener("pointercancel", stopDragging); viewport.addEventListener("dblclick", resetView); plusButton.addEventListener("click", function () { changeScale(1.25); }); minusButton.addEventListener("click", function () { changeScale(0.8); }); resetButton.addEventListener("click", resetView); closeButton.addEventListener("click", function () { dialog.close(); }); dialog.addEventListener("click", function (event) {
    if (event.target === dialog) dialog.close();
}); });
</script>

# Interpretation guardrails

- Mortality predictors are cumulative percentages over each complete condition history; they are not annualized rates in these preliminary models.
- Cumulative site CWD is TerraClimate `def` exposure at the FIA site, not the CWD community-niche response.
- The raw figures are descriptive. Use the model tables and marginal-effect figures when discussing covariate-adjusted model results.
- Saplings, adults, and the combined sapling-plus-adult forest community are shown as three separate model columns and plot panels.
- Seedlings are excluded because their microplot sampling does not reliably align with the condition-level disturbance predictors.
